```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
### Health Access in Haiti
# P. Pokharel, Sep. 2013

dat_dir <- "~/Code/Haiti_Hospital_Access/"
setwd(dat_dir)

### Installing packages
# install.packages(c("raster", "rgdal", "rasterVis"))
### Load packages
require(raster)
require(rgdal)
require(rasterVis)
```

The datasets we have access to are a population grid estimate from 2003, and a list of health facilities from before the 2010 earthquake, which seems to have primarily been collected in 2005.

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
## Load health facility data, and population grid
health_df <- read.csv("data/HealthFacilities.csv", stringsAsFactors=F)
  # there is a single row where the lat/lng is off from the rest of dataset; remove it
  health_df <- health_df[-which(health_df$longitude > -10),]
#popgrid <- readGDAL("data/2003_pop_estimates.tif")
popraster <- raster("data/2003_pop_estimates.tif")

## Re-project to the same co-ordinate system
health_sp <- health_df
coordinates(health_sp) <- ~longitude+latitude
proj4string(health_sp) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
health_sp <- spTransform(health_sp, popraster@crs)
health_df <- as.data.frame(health_sp) # in case we want to poke around a normal df

## To make things faster, run operations on an aggregated raster
## which is NA'ed out in areas with no people
## Scale: one pixel used to be 100mx100m, now it is 10kmx10km
popagg <- aggregate(popraster, fact=10, fun=sum)
values(popagg) <- replace(values(popagg), values(popagg)<1, NA)
```

Here, roughly, is what a coarse view of (log) population in Haiti looks like:

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
blank_theme <- function() {
  theme(axis.text=element_blank(), axis.ticks=element_blank(), panel.grid=element_blank(), axis.title=element_blank())
}
poplayer <- gplot(popagg) + geom_tile(aes(fill=value)) + scale_fill_continuous(trans='log10') + blank_theme() + labs(fill="Log Population\n(per 10km^2 cell)")
print(poplayer)
```

The objective here is to think about population and the health facilities. Lets make a quick plot, to send our minds in the right place; red is health facilities here:
```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
poplayer + geom_point(data=health_df, aes(x=longitude, y=latitude), color='red', alpha=0.5)
```


We don't get a very clear picture of things there. Next step is to calculate distances from populated points to facilities.
```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
## Find distances from health facility
distraster <- distanceFromPoints(popagg, health_sp, 
  filename='data/distanceRaster.tif', format='GTiff', overwrite=TRUE)
distraster <- mask(distraster, popagg)
```

Finally, lets plot it! 
Bluer the point, farther away from a health facility. Red points are health facilities, overlaid on top:
```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
distlayer <- gplot(distraster) + geom_tile(aes(fill=value)) +
  blank_theme() + labs(fill="Distance to Health Facility")
distlayer + geom_point(data=health_df, aes(x=longitude, y=latitude), color='red', alpha=0.5)
```


```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
require(plyr)
health_df <- plyr::rename(health_df,c("latitude"="y", "longitude"="x"))
health_df$row <- row.names(health_df)
health_df <- subset(health_df, select=c("x", "y", "row", "Type"))

# For each population point, identify the closest health clinic
popagg2 <- raster::aggregate(popagg)
popPoints <- as.data.frame(rasterToPoints(popagg2))
require(RSAGA)
nearestH <- pick.from.points(popPoints, health_df, radius=2E10, pick="row")
```

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
merged <- merge(nearestH, health_df, by="row", 
                suffixes=c(".pop", ".health"))
merged$pID <- row.names(merged)
imerged <- idata.frame(merged)
mergedPlotDF <- ddply(imerged, .(pID), function(df) { 
  data.frame(
    x=c(df$x.pop, df$x.health), 
    y=c(df$y.pop, df$y.health), 
    group=c(df[1,'pID'], df[1,'pID']),
    Population=c(df[1,'X2003_pop_estimates'],df[1,'X2003_pop_estimates'])
)})
```


```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12}
ggplot(mergedPlotDF, aes(x=x, y=y, group=group)) + geom_line() + blank_theme()
```

And if we want to look at population as well, here is what we see:

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12}
ggplot(mergedPlotDF, aes(x=x, y=y, group=group, color=Population)) + geom_line() + blank_theme() + scale_color_continuous(trans='log10')
```

We can also look at the number of people served by a given health facility (roughly), and look at that according to the type of the health facility. For example, small clinics that are serving huge populations may be due for an upgrade. Unfortunately, the typology in our dataset isn't great, but nevertheless, here is a broad brush picture:

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
popServed <- ddply(imerged, .(row, Type), function(df) { 
  data.frame(
    popServed = sum(df$X2003_pop_estimates), 
    x=df[1,'x.health'], 
    y=df[1,'y.health']
)})
ggplot(data=popServed, aes(x=x, y=y, size=popServed, color=Type)) + geom_point() + blank_theme()
```