---
title       : Visualizing access to health in Haiti
subtitle    : 
author      : Prabhas Pokharel, Modi Research Group
job         : 
framework   : io2012      # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
## Visualizing access to health in Haiti

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
### Health Access in Haiti
# P. Pokharel, Sep. 2013

dat_dir <- "~/Code/Haiti_Hospital_Access/"
setwd(dat_dir)

### Installing packages
# install.packages(c("raster", "rgdal", "rasterVis", "ggplot2", 
# "plyr", "doBy"))
### Load packages
require(raster)
require(rgdal)
require(rasterVis)
require(ggplot2)
```

In this presentation, we will look at visualizations of access to health facilities in Haiti.


The datasets we are working with are a population estimate, 2003, from the US Census Bureau. The resolution of this dataset is 100m^2 per cell/pixel, but the analysis in this presentation will be done at an aggregation of 1km^2 per pixel.

The health facility dataset is from [GeoCommons](http://geocommons.com/overlays/21222).

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE, results='hide'}
## Load health facility data, and population grid
health_df <- read.csv("data/HealthFacilities.csv", stringsAsFactors=F)
  # there is a single row where the lat/lng is off from the rest of dataset; remove it
  health_df <- health_df[-which(health_df$longitude > -10),]
  # there is no consistent id, so lets use the row.name
  health_df$hID <- row.names(health_df)
#popgrid <- readGDAL("data/2003_pop_estimates.tif")
popraster <- raster("data/2003_pop_estimates.tif")

## Re-project to the same co-ordinate system
health_sp <- health_df
coordinates(health_sp) <- ~longitude+latitude
wgs84 <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
utm <- popraster@crs
proj4string(health_sp) <- wgs84
health_sp <- spTransform(health_sp, utm)
health_df_utm <- as.data.frame(health_sp)

## Read the Haiti admin 3 shapefile and re-project data to utm
require(maptools); require(gpclib); gpclibPermit()
admin_regions <- spTransform(readOGR('data/', layer='Haiti_Sections_Final_WGS84'), utm)
admin_regions_DF <- fortify(admin_regions, region='ID_COMMUNE')

## To make things faster, run operations on an aggregated raster
## which is NA'ed out in areas with no people
## Scale: one pixel used to be 100mx100m, now it will be 1kmx1km
popagg <- aggregate(popraster, fact=10, fun=sum)
values(popagg) <- replace(values(popagg), values(popagg)<1, NA)
```

---
## Haiti's population

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
blank_theme <- function() {
  theme(axis.text=element_blank(), axis.ticks=element_blank(), 
        panel.grid=element_blank(), axis.title=element_blank())
}
haiti_overlay <- function() {
  geom_path(data=admin_regions_DF, aes(x=long, y=lat, group=group), color='#88888888')
}
require(ggmap)
poplayer <- gplot(popagg) + geom_tile(aes(fill=value)) + 
  scale_fill_gradient(low='white', high='darkblue', trans='log10', na.value='#ffffff00') + 
  blank_theme() + labs(fill="Log Population\n(per 1km^2 cell)")
poplayer + haiti_overlay()
```
The total population represented is 8.4M; Haiti's population has grown to 10M+ in 2013.
We will be working at a resolution of 1km^2 per cell; shown above.

---
## Haiti's population and health facilities

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
poplayer + 
    geom_point(data=health_df_utm, aes(x=longitude, y=latitude), 
               color='darkred', shape='+', size=4.5) +
    haiti_overlay()
```

Here, health facilities (in red) have been overlaid on top of the population layer.
The dataset, from GeoCommons, was primarily collected in 2005 and revised in 2009.

--- .segue .dark

## 1. Where are people far away from health facilities?

---

Below, the redder the pixel, the farther the people living there have to travel to a health facility.

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
## Find distances from health facility
distraster <- distanceFromPoints(popagg, health_sp, 
  filename='data/distanceRaster.tif', format='GTiff', overwrite=TRUE)
distraster <- mask(distraster, popagg)
```

```{r echo=F, warning=F, message=F, comment=NA, fig.height=6, fig.width=10, cache=TRUE}
distlayer <- gplot(distraster) + geom_tile(aes(fill=value)) +
  blank_theme() + labs(fill="Distance to\nHealth Facility") +
  scale_fill_gradient2(low='white', mid='yellow', high='red', na.value='#ffffff00')
distlayer + haiti_overlay()
```

Methodology: straight line distance in m (on UTM projection).

---

**A different view:** The longer the line, the farther people have to travel to the nearest health facility--lines connection population points to health facilities. (Inspired by [[1]](http://flowingdata.com/2013/08/27/in-search-of-food-deserts/))

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
require(plyr)
health_df_utm <- plyr::rename(health_df_utm,c("latitude"="y", "longitude"="x"))
health_df_utm <- subset(health_df_utm, select=c("x", "y", "hID", "Categorie"))

# For each population point, identify the closest health clinic
popagg2 <- raster::aggregate(popagg)
popPoints <- as.data.frame(rasterToPoints(popagg2))
require(RSAGA)
nearestH <- pick.from.points(popPoints, health_df_utm, radius=2E10, pick="hID")
```

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
# create a merge dataset; each row is a population point + nearest health facility
merged <- merge(nearestH, health_df_utm, by="hID", 
                suffixes=c(".pop", ".health"))
# also pull lat / long in
merged <- merge(merged, health_df[c("hID", "longitude", "latitude")], by="hID")
merged$pID <- row.names(merged)
merged$distance <- sqrt((merged$x.pop - merged$x.health)^2 + (merged$y.pop - merged$y.pop)^2)
imerged <- idata.frame(merged)

# create a data frame for plotting; instead of health facility and nearest data frame
# on single row, separate them out, but match them by "group" name
mergedPlotDF <- ddply(imerged, .(pID), function(df) { 
  data.frame(
    x=c(df$x.pop, df$x.health), 
    y=c(df$y.pop, df$y.health),
    group=c(df[1,'pID'], df[1,'pID']),
    Population=c(df[1,'X2003_pop_estimates'],df[1,'X2003_pop_estimates']),
    type=c("Population Point", "Health Facility")
)})
```

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=12, cache=TRUE}
require(ggplot2); require(OpenStreetMap)
mapprojected <- readRDS('data/HaitiBaseMapPlain_utm.rds')
autoplot(mapprojected)  + blank_theme() + 
    geom_line(data=mergedPlotDF, aes(x=x, y=y, group=group))
```

--- .segue .dark

## 2. Where are there large populations far away from health facilities?

---

On this map, population is represented by color and lines are the same as before. Deeper purple lines indicate bigger populations.

```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=13, cache=TRUE}
mergedPlotDF$Population.Bin <- 
    cut(mergedPlotDF$Population, include.lowest=T,
        breaks=c(1,100,1000,10000,100000),
        labels=c("Below 100", "100-1,000", "1,000-10,000", "> 10,000"))
autoplot(mapprojected)  + blank_theme() +  
    geom_line(data=mergedPlotDF, aes(x=x, y=y, group=group, 
                                     color=Population.Bin)) + 
    scale_color_brewer(palette="PuRd")
```

--- .segue .dark


```{r echo=F, warning=F, message=F, comment=NA, fig.height=8, fig.width=13, cache=TRUE}
```

---

## And next ...

 * These maps are the first steps of analysis, more sophisticated algorithsm can dive deeper, including into recommendations.

 * The pre-requisite is good data; for garbage in, garbage out.

 * The analysis should be made interactive for local-level users.

 * And finally, the analysis, however sophisticated, cannot exist in a vacuum. It needs to be interpreted and understood by those who actually know on-the-ground realities.
 
We are working on all of these problems at the [Sustainable Engineering Lab](http://modi.mech.columbia.edu).

--- .centered
## Thank You

Prabhas Pokharel

[Sustainable Engineering Lab](http://modi.mech.columbia.edu)

Earth Institute

Columbia University
